---
title: 'CreelCatch: a package to integrate creel surveys over space and time'
author: "Matthew Robertson"
date: "30/09/2022"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(TMB)
library(CreelCatch)
library(ggplot2)
```

# Overview

CreelCatch is a package designed to estimate creel surveys over space and time by using socio-ecological covariates and sampling design random effects to estimate fishing effort and then using that fishing effort to estimate catch. 

# Installation

Install the [devtools](https://cran.r-project.org/web/packages/devtools/index.html) package and run:

```{r, eval=FALSE}
devtools::install_github(MatthewRobertson2452\CreelCatch)
library(CreelCatch)
```

If you are having problems with installation, you can install the package locally as a ZIP file by clicking the Code menu and "download ZIP" from the [github page](https://github.com/MatthewRobertson2452/CreelCatch). You can then extract the folder in a local directory while recording the directory name (which I will reference as download_dir). To install, then use
```{r, eval=FALSE}
devtools::install_local(path=download_dir, dependencies=FALSE)
library(CreelCatch)
```

# Basic Example

The examples will use the provided creel survey dataset and describe multiple model formulations that can be used to estimate catch and effort.

To load the provided dataset, 

```{r}
data("example_dat")
```

This data contains the same information that is described in the paper (Robertson et al. In prep), and information detailing what each column contains can be examined in detail with

```{r, eval=FALSE}
?example_dat
```

Creating the objects needed to run the model depends on then use of the `model_prep` function and the description of desired model formulation within that function. To start this example, I will show a simple model formulation without random effects or covariates.

To run the most basic version of the model, you only need to specify the creel estimates of catch and effort. We found the model provided the best fit when examining both of these on then log-scale, however, this transformation is not forced by the `model_prep` function.

```{r}
output<-model_prep(catch = log(example_dat$Catch_per_day), effort = log(example_dat$Effort_per_day))
```

The `model_prep` function provides multiple outputs in a named list that then need to be input into TMB to run the model, these include the list of data objects (`tmb.data`), parameters (`parameters`), what parameters should be treated as random effects (`rname`), and what parameters do not need to be estimated (`map`).

These outputs can then be used in the following TMB functions to prepare and run the model.

```{r, results="hide", message=FALSE}
obj <- TMB::MakeADFun(data = c(model = "CreelCatch", # which model to use
                                output$tmb.data),
                       parameters = output$parameters,
                       map=output$map,random = output$rname,
                       DLL = "CreelCatch_TMBExports", silent=TRUE)

opt<-nlminb(obj$par,obj$fn,obj$gr,
            control = list(trace=10,eval.max=2000,iter.max=1000))
```

The model output data will be stored in `opt` and we can initially determine model convergence and fit information by examining the gradient, the negative log-likelihood and AIC and BIC,

```{r}
####Largest Gradient (smaller is better)
max(obj$gr())

####Negative log-likelihood
opt$objective

#####AIC
2*opt$objective+ 2*length(opt$par)

#####BIC
2*opt$objective + log(length(output$tmb.data$log_C))*length(obj$par)
```

Parameter estimates and their associated uncertainty can then be examined using the report object (estimates), and sdreport object (estimates and uncertainty),

```{r}
rep <- obj$report()
sdrep<-TMB::sdreport(obj)
```

To manually examine parameter estimates you can index the parameter of interest as follows (e.g. for catchability),

```{r}
rep$q_dev

#indexing the sdreport object for parameter estimates
val_names<-names(sdrep$value)
ind<-val_names=="q_dev"
sdrep$value[ind]

#indexing the sdreport object for uncertainty estimates
sdrep$sd[ind]
```

I have also added functionality to easily plot the estimated relationship between catch and effort by using the function `plot_cpue`

```{r}
plot_cpue(rep=rep, sdrep=sdrep, tmb.data=output$tmb.data)
```

# Random Effects and Covariates Example

While this basic example is useful to get a basic understanding of how the model works, most uses of the model will seek to explore how the model estimates can be improved by accounting for survey design characteristics with random effects or by accounting for variability in fishing effort with socio-ecological covariates. This example will show how both of these can be included.

Our best model formulation came when the effort intercept varied by region and where effort was modified by waterbody log(area). To include these two effects, we need to slightly modify the `model_prep` function. The function includes various "switches" that tell the model to turn on different effects. These include `E_intercept` to treat the effort intercept as a random effect, `C_intercept` to estimate the catch intercept and treat it as a random effect, `catchability` to treat catchability as a random effect, `n_covars` to tell the model how many covariates you would like to include, and `gl_switch` to indicate whether to differentially account for the waterbody area of Great Lakes surveys (covered in the following example).

Turning on switches sometimes requires inputting additionally data as well. For example, turning on a random effect requires inputting data describing that random effect. Random effect data should be integers to denote separate groups, where the smallest value is 0. 

Here, we told the model to estimate the effort intercept as a random effect by state and to use 1 covariate (specifically log waterbody area) to describe effort.

```{r}
output<-model_prep(catch = log(example_dat$Catch_per_day), effort = log(example_dat$Effort_per_day),
                   E_intercept = 1, random_effect = example_dat$istate, n_covars = 1, 
                   first_covar = log(example_dat$Area))
```

We can then run the model the same as before,

```{r}
obj <- TMB::MakeADFun(data = c(model = "CreelCatch", # which model to use
                                output$tmb.data),
                       parameters = output$parameters,
                       map=output$map,random = output$rname,
                       DLL = "CreelCatch_TMBExports", silent=TRUE)

opt<-nlminb(obj$par,obj$fn,obj$gr,
            control = list(trace=10,eval.max=2000,iter.max=1000))

#####AIC
2*opt$objective+ 2*length(opt$par)

#####BIC
2*opt$objective + log(length(output$tmb.data$log_C))*length(obj$par)

rep <- obj$report()
sdrep<-TMB::sdreport(obj)
```

I have also included functionality to be able to easily plot estimates of the random effects and covariates.

For the random effects, you need to input the object created from `model_prep`, the sdrep object, and then the names associated with each level of the random effect and the name describing what the random effect was. The model will show the random effect point estimates and associated 95% confidence intervals, but does not account for multiple comparisons.

```{r}
plot_re(output=output, sdrep=sdrep, names=c("Michigan","Florida","Connecticut","Kansas","South Dakota"), re_name="State")
```

For covariates, you need to input model output, the object from `model_prep`, the number of covariates that were used, and the names of those covariates.

```{r}
plot_covars(rep=rep, sdrep=sdrep, output=output, n_covar=1, covar_name=c("Waterbody Area"))
```
  
# Accounting for the Great Lakes Example

During our model testing, we identified different relationships between fishing effort and area for creel surveys conducted on the Great Lakes. To account for this, our model has the capability of separately estimating these relationships if the Great Lakes surveys are indexed. However, including this functionality can add some complexity to model setup, so I will provide a brief example here.

Since estimating waterbody area is itself a covariate, n_covar must account for this. However, because the Great Lakes waterbody area is entered into the model as a unique variable, if additional covariates are included they need to be input into `model_prep` as covar-1. For example, if we wanted to examine waterbody area with separate slopes for Great Lakes waterbodies and the effect of median county age our `model_prep` would be,


```{r}
output<-model_prep(catch = log(example_dat$Catch_per_day), effort = log(example_dat$Effort_per_day),
                   E_intercept = 1, random_effect = example_dat$istate,
                   n_covars = 2, first_covar = log(example_dat$Age),
                   gl_switch = 1, gl_area=log(example_dat$Area), iGL=example_dat$GL)
```

Where gl_area is the input for waterbody area and iGL is the index of whether surveys occurred on the Great Lakes or not.

Once this step is done, running the model and plotting is the same as before but with a slight modification for plotting covariates. When plotting covariates you only need to enter the name of covariates in addition to waterbody area and you need to indicate that gl_switch=1.

```{r, echo=FALSE}
obj <- TMB::MakeADFun(data = c(model = "CreelCatch", # which model to use
                                output$tmb.data),
                       parameters = output$parameters,
                       map=output$map,random = output$rname,
                       DLL = "CreelCatch_TMBExports", silent=TRUE)

opt<-nlminb(obj$par,obj$fn,obj$gr,
            control = list(trace=10,eval.max=2000,iter.max=1000))

#####AIC
2*opt$objective+ 2*length(opt$par)

#####BIC
2*opt$objective + log(length(output$tmb.data$log_C))*length(obj$par)

rep <- obj$report()
sdrep<-TMB::sdreport(obj)
```



```{r}
plot_covars(rep=rep, sdrep=sdrep, output=output, n_covar=2, covar_name=c("Age"), gl_switch=1)
```
